# Missing values
```{r}
library(tidyverse)
library(patchwork)
library(reshape2)


missing_plot <- function(data, percent=FALSE)
  {
    missing_patterns <- data.frame(is.na(data)) %>%
      group_by_all() %>%
      count(name = "count", sort = TRUE) %>%
      ungroup()
    if(percent)
      {
        missing_patterns <- missing_patterns %>%
        mutate(count=count/sum(count)*100)
      }

sort_data <- names(sort(colSums(is.na(data)), decreasing = TRUE))
    
dataNew <- missing_patterns %>% 
  subset(select=-count)

id <- which(rowSums(dataNew)==0)
row <- nrow(dataNew)
col <- ncol(dataNew)

dataNew <- dataNew %>% 
  mutate(ID = rownames(.)) %>%
  melt(id.vars=c("ID")) %>%
  mutate(missing = ifelse(ID == id,2,ifelse(value == "TRUE", 1, 0)))
  
main_ <- ggplot(dataNew, aes(x=factor(variable,levels=sort_data), 
                                 y=factor(ID,rev(unique(ID))))) + 
  geom_tile(aes(fill=factor(missing)),color = "white") + 
  scale_fill_manual(values = alpha(c("grey", "blueviolet","grey"), c(.4,.4,0.9))) + 
  geom_text(aes(x=(1+col)/2,y=row-id+1), label = "complete cases") + 
  scale_x_discrete(label = function(x) abbreviate(x, minlength =5))+
  xlab('variable') + 
  ylab('missing pattern') + 
  guides(fill='none') + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
col_ <- ggplot(missing_patterns, aes(x=factor(rownames(missing_patterns), rev(rownames(missing_patterns))), y=count)) + 
  geom_bar(stat='identity', fill=ifelse(rownames(missing_patterns)==id, 'cornflowerblue', alpha('cornflowerblue',0.5))) + 
  xlab('') + 
  ylab(ifelse(percent,'%rows','row count')) + 
  theme_bw() + 
  theme(panel.grid.major.y=element_blank()) + 
  coord_flip()
  
  if(percent)
  {
    col_<-col_+ylim(0,100)
  }
  
DfThree <- data.frame(colSums(is.na(data)))
  colnames(DfThree) <- 'count'
  
    if(percent)
      {
        DfThree <- DfThree%>%mutate(count=count/nrow(data)*100)
      }
  
  row_ <- ggplot(DfThree,aes(x=factor(rownames(DfThree),
                                    levels=sort_data), y=count)) +
      geom_bar(stat='identity',fill=alpha('cornflowerblue',0.5)) +
      ggtitle('missing value patterns') +
      theme_bw() +
      theme(panel.grid.major.x=element_blank()) +
      scale_x_discrete(label = function(x) abbreviate(x, minlength =5))+
      xlab('') +
      ylab(ifelse(percent,'%rows missing','num rows missing')) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
  if(percent)
    {
      row_<-row_ + ylim(0,100)
    }
  
  layout_ <- "
              111#
              2223
              2223
              2223
                    "
  row_+main_+col_+plot_layout(design=layout_)
}
```

```{r}
data <- read.csv(file ='./athlete_events.csv')
missing_plot(data, TRUE)
```

```{r}
missing_plot(data, FALSE)
```

## Observations
- In an event, exactly three people win medals, ie. Gold, Silver and Bronze, as a result around 75% values in the medal's column are missing.
- In some sporting events, the weight and height are not necessarily measured. Consequently, around 25% values are missing as well. Upon inspection, we observe that weight and height measurements are missing together in most of the cases.
- Since the other columns like Name, City, Sport, Sex, Team, NOC, Games, Year, and Season are very important, there are no missing values in these columns.
- Surprisingly, there are some missing values in the Age column as well. By simple observation, we see that these rows belong to players (for instance, S. Abdul Hamid) that participated during the Olympic Games from 1920s-1950s. We hypothesize that these records must have been lost as a result they're missing.